An algorithm for the optimization of finite element integration loops

### 2016-12-23 TOMS_Paper/reviewers_answers.txt

Referee: 1

Comments to the Author This paper is overall well written. Examples are very
illustrative and help understand the article. My only concern is about how
authors relates their theortical results with numerical performance examples.
The main focus of this paper is about transformations that reduces FLOPs.
Authors present elimination and reduction pattern of code and show that there is
an optimal loop transformation that reduces FLOPs. However, the numeric
performance results is about "speed-up" not about FLOPs. Also, the speed-up is
more significant when the polynomial degree is higher. This confuses me either
vectorization improves the performance or actual FLOPs decreases. My suggestions
are as follows: 1. It would be better to include actual transformed code and
original code in the appendix for all computing model (mass, helmholtz,
elasticity, hyper elasticity). At least I wonder what makes hugh speed up in the
hyper elastic problem compared to usual mass matrix.  2. Please include actual
FLOP measurements of input code and transformed code and show some correlation
with performance.

Referee: 2

Comments to the Author

Report on “An algorithm for the optimization of finite element integration
loops”

The paper is well written, interesting does clearly merit publication.  The
usage of advanced compiler technology as demonstrated here is clearly novel and
it brings accelerated performance.

There is one issue that is not covered in the analysis and only briefly
mentioned in the general discussion: the transfer from memory. I understand that
it complicates the analysis but the least it should be quantified in the
numerical experiments. As was shown by Alnæs and Mardal in 2010, TOMS, the
transfer is significant. This part of the story should also be told as I guess
the most common user, low-order elements with simple equation, will not notice
any benefit. This has been an untold story in several FEniCS papers. It is to a
certain extent discussed already in the paper but it should also be quantified.

Referee: 3

Comments to the Author The FEM is widely used in scientific computing and the
integration to compute the elements (local element matrices)  can often be time
consuming. This manuscript is about optimizing the finite element integration
loops and the number of flops therein. The authors propose an approach that is
shown to be locally optimal. An implementation (in COFFEE) is compared against
other methods and speedups over the default generated code is demonstrated in
several cases.

Overall this is a well written and interesting paper, even to non-experts.
However, as a non-expert, I have several concerns that I think should be
addressed:

1) The focus is on minimizing flops, but on modern architectures, memory
accesses are typically more expensive and flops are "almost free". Although
clearly of theoretical interest, it would be nice to justify that this
optimization also is important to reduce run time.

2) All experiments and results are run in serial on single core. In practice,
most computers today are multicore (or manycore) so some discussion about
extensions to multithreaded case would be appropriate. I recognize a parallel
implementation is beyond the scope of the current paper.

3) Results: The authors present several charts with speedups on different
problems, which gives detailed information. However, it would be nice to add a
figure with aggregate results, such as a scatter plot of all instances or a
performance profile.

4) The manuscript is already quite long so if possible, do not increase the
total length (i.e. try find things to tighten or cut).Referee: 1

Comments to the Author This paper is overall well written. Examples are very
illustrative and help understand the article. My only concern is about how
authors relates their theortical results with numerical performance examples.
The main focus of this paper is about transformations that reduces FLOPs.
Authors present elimination and reduction pattern of code and show that there is
an optimal loop transformation that reduces FLOPs. However, the numeric
performance results is about "speed-up" not about FLOPs. Also, the speed-up is
more significant when the polynomial degree is higher. This confuses me either
vectorization improves the performance or actual FLOPs decreases. My suggestions
are as follows: 1. It would be better to include actual transformed code and
original code in the appendix for all computing model (mass, helmholtz,
elasticity, hyper elasticity). At least I wonder what makes hugh speed up in the
hyper elastic problem compared to usual mass matrix.  2. Please include actual
FLOP measurements of input code and transformed code and show some correlation
with performance.

Referee: 2

Comments to the Author

Report on “An algorithm for the optimization of finite element integration
loops”

The paper is well written, interesting does clearly merit publication.  The
usage of advanced compiler technology as demonstrated here is clearly novel and
it brings accelerated performance.

There is one issue that is not covered in the analysis and only briefly
mentioned in the general discussion: the transfer from memory. I understand that
it complicates the analysis but the least it should be quantified in the
numerical experiments. As was shown by Alnæs and Mardal in 2010, TOMS, the
transfer is significant. This part of the story should also be told as I guess
the most common user, low-order elements with simple equation, will not notice
any benefit. This has been an untold story in several FEniCS papers. It is to a
certain extent discussed already in the paper but it should also be quantified.

Referee: 3

Comments to the Author The FEM is widely used in scientific computing and the
integration to compute the elements (local element matrices)  can often be time
consuming. This manuscript is about optimizing the finite element integration
loops and the number of flops therein. The authors propose an approach that is
shown to be locally optimal. An implementation (in COFFEE) is compared against
other methods and speedups over the default generated code is demonstrated in
several cases.

Overall this is a well written and interesting paper, even to non-experts.
However, as a non-expert, I have several concerns that I think should be
addressed:

1) The focus is on minimizing flops, but on modern architectures, memory
accesses are typically more expensive and flops are "almost free". Although
clearly of theoretical interest, it would be nice to justify that this
optimization also is important to reduce run time.

2) All experiments and results are run in serial on single core. In practice,
most computers today are multicore (or manycore) so some discussion about
extensions to multithreaded case would be appropriate. I recognize a parallel
implementation is beyond the scope of the current paper.

3) Results: The authors present several charts with speedups on different
problems, which gives detailed information. However, it would be nice to add a
figure with aggregate results, such as a scatter plot of all instances or a
performance profile.

4) The manuscript is already quite long so if possible, do not increase the
total length (i.e. try find things to tighten or cut).Referee: 1

Comments to the Author This paper is overall well written. Examples are very
illustrative and help understand the article. My only concern is about how
authors relates their theortical results with numerical performance examples.
The main focus of this paper is about transformations that reduces FLOPs.
Authors present elimination and reduction pattern of code and show that there is
an optimal loop transformation that reduces FLOPs. However, the numeric
performance results is about "speed-up" not about FLOPs. Also, the speed-up is
more significant when the polynomial degree is higher. This confuses me either
vectorization improves the performance or actual FLOPs decreases. My suggestions
are as follows: 1. It would be better to include actual transformed code and
original code in the appendix for all computing model (mass, helmholtz,
elasticity, hyper elasticity). At least I wonder what makes hugh speed up in the
hyper elastic problem compared to usual mass matrix.  2. Please include actual
FLOP measurements of input code and transformed code and show some correlation
with performance.

Referee: 2

Comments to the Author

Report on “An algorithm for the optimization of finite element integration
loops”

The paper is well written, interesting does clearly merit publication.  The
usage of advanced compiler technology as demonstrated here is clearly novel and
it brings accelerated performance.

There is one issue that is not covered in the analysis and only briefly
mentioned in the general discussion: the transfer from memory. I understand that
it complicates the analysis but the least it should be quantified in the
numerical experiments. As was shown by Alnæs and Mardal in 2010, TOMS, the
transfer is significant. This part of the story should also be told as I guess
the most common user, low-order elements with simple equation, will not notice
any benefit. This has been an untold story in several FEniCS papers. It is to a
certain extent discussed already in the paper but it should also be quantified.

Referee: 3

Comments to the Author The FEM is widely used in scientific computing and the
integration to compute the elements (local element matrices)  can often be time
consuming. This manuscript is about optimizing the finite element integration
loops and the number of flops therein. The authors propose an approach that is
shown to be locally optimal. An implementation (in COFFEE) is compared against
other methods and speedups over the default generated code is demonstrated in
several cases.

Overall this is a well written and interesting paper, even to non-experts.
However, as a non-expert, I have several concerns that I think should be
addressed:

1) The focus is on minimizing flops, but on modern architectures, memory
accesses are typically more expensive and flops are "almost free". Although
clearly of theoretical interest, it would be nice to justify that this
optimization also is important to reduce run time.

2) All experiments and results are run in serial on single core. In practice,
most computers today are multicore (or manycore) so some discussion about
extensions to multithreaded case would be appropriate. I recognize a parallel
implementation is beyond the scope of the current paper.

3) Results: The authors present several charts with speedups on different
problems, which gives detailed information. However, it would be nice to add a
figure with aggregate results, such as a scatter plot of all instances or a
performance profile.

4) The manuscript is already quite long so if possible, do not increase the
total length (i.e. try find things to tighten or cut).

### 2016-12-23 TOMS_Paper/reviewers_response.txt

First of all, we would like to thank the referees for their insightful comments,
which helped improving the paper. In this new version, we have attempted to
address all of the referees' concerns. We explain how the paper was changed
below, by considering each of the requests/observations/questions raised by the
three referees.

Referee: 1
==========

| This paper is overall well written. Examples are very illustrative and help
| understand the article. My only concern is about how authors relates their
| theortical results with numerical performance examples. [...]

| My suggestions are as follows:
| 1. It would be better to include actual transformed code and original code in
| the appendix for all computing model (mass, helmholtz, elasticity, hyper
| elasticity). At least I wonder what makes hugh speed up in the hyper elastic
| problem compared to usual mass matrix.

The generated code has been uploaded to GitHub and given a DOI through Zenodo.
This is properly documented in the paper (see Section 7.1 - Experimental setup).
For reasons of space, it simply was not possible to append code in an appendix.

| 2. Please include actual FLOP measurements of input code and transformed code
| and show some correlation with performance.

There is a whole new section treating this fundamental aspect (7.8 - On the
correlation between operation count and performance). Further, a new figure was
added (Fig. 11): this clearly shows the relationship between the achieved
speed-ups and FLOPs reductions. We believe that this figure is particularly
useful, because it shows how well the system presented in the paper scales well
with the complexity of operators. It also provides a nice summary of the
performance improvements achieved. In addition to all this, Sections 7.4 through
7.7 (Mass, Helmholtz, ...) have been extended and now discuss the impact of the
reduced operation counts in a set of relevant test cases.

Referee: 2
==========

| The paper is well written, interesting does clearly merit publication.
| The usage of advanced compiler technology as demonstrated here is clearly
| novel and it brings accelerated performance.

| There is one issue that is not covered in the analysis and only briefly
| mentioned in the general discussion: the transfer from memory. I understand
| that it complicates the analysis but the least it should be quantified in
| the numerical experiments [...] It is to a certain extent discussed already
| in the paper but it should also be quantified.

Sections 7.4 through 7.7 now discuss, to some extent, the memory transfer cost.
We did that only for a small set of (relevant) test cases. We believe that it is
very difficult to thoroughly study this problem in a paper like ours, based on a
high dimensional experimentation space. This would probably require its own
study in a separate paper, introducing specific metrics and showing different
angles of the performance achieved. For all these reasons, we believe that it is
just simpler to refer to a paper that, in fact, has performed an analysis
analogous to what the referee is suggesting: [Bercea et al.  2016].  This is
referenced in Section 7.2, along with a high level discussion of the memory
transfer impact.  Notably important, the Bercea paper is based upon the same
Firedrake technology of our article (in fact, we are amongst its co-authors).

Referee: 3
==========

| Overall this is a well written and interesting paper, even to non-experts.
| However, as a non-expert, I have several concerns that I think should be
| addressed:

| 1) The focus is on minimizing flops, but on modern architectures, memory
| accesses are typically more expensive and flops are "almost free". Although
| clearly of theoretical interest, it would be nice to justify that this
| optimization also is important to reduce run time.

Figures 7, 8, 9, 10 show execution time speed-ups. Hence, we are already showing
that our work contributes to actually reducing run-times.

| 2) All experiments and results are run in serial on single core. In practice,
| most computers today are multicore (or manycore) so some discussion about
| extensions to multithreaded case would be appropriate. I recognize a parallel
| implementation is beyond the scope of the current paper.

We have added one paragraph in Section 9 and added a few sentences throughout
the text about this matter. Basically, in Firedrake parallelism is achieved at
the mesh level (by distributing mesh partitions to processes), while the
execution of a local assembly kernel is sequential. Clearly this might need to
be revisited in the case of higher order methods (as we would end up with big
working sets even for executing just a single element). This is discussed in
Section 9.

| 3) Results: The authors present several charts with speedups on different
| problems, which gives detailed information. However, it would be nice to add a
| figure with aggregate results, such as a scatter plot of all instances or a
| performance profile.

We agree with the referee about the usefulness of such a chart. We have added a
scatter plot (Fig. 11) as well as a section explaining what information we can
extrapolate from it.

| 4) The manuscript is already quite long so if possible, do not increase the
| total length (i.e. try find things to tighten or cut).

We tried our best, but we feel that we cannot really stay within 27 pages
without losing essential information. We observe, however, that other TOMS
papers went even beyond the 27 pages.

### 2016-12-30 TOMS_Paper/assembly.tex
minor fixes

We evaluate the speed-ups achieved by a wide variety of transformation systems over the ``original'' code produced by the FEniCS Form Compiler (i.e., no optimizations applied). We analyze the following transformation systems:
\begin{description}
\item[quad] Optimized quadrature mode. Work presented in~\citeN{quadrature1}, implemented in  in the FEniCS Form Compiler. 
\item[tens] Tensor contraction mode. Work presented in~\citeN{FFC-TC}, implemented in the FEniCS Form Compiler.
\item[auto] Automatic choice between \texttt{tens} and \texttt{quad} driven by heuristic (detailed in~\citeN{Fenics} and summarized in Section~\ref{sec:qualitative}). Implemented in the FEniCS Form Compiler.
\item[ufls] UFLACS, a novel back-end for the FEniCS Form Compiler whose primary goals are improved code generation and execution times.
\item[cfO1] Generalized loop-invariant code motion. Work presented in~\citeN{Luporini}, implemented in COFFEE.
\item[cfO2] Optimal loop nest synthesis with handling of block-sparse tables. Work presented in this article, implemented in COFFEE.
\end{description}

@@

We report the results of our experiments in Figures~\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity} as three-dimensional plots. The axes represent $q$, $\mathrm{nf}$, and code transformation system. We show one subplot for each problem instance ${\langle} \textrm{form}, \mathrm{nf}, q {\rangle}$, with the code transformation system varying within each subplot. The best variant for each problem instance is given by the tallest bar, which indicates the maximum speed-up over non-transformed code. We note that if a bar or a subplot are missing, then the form compiler failed to generate code because it either exceeded the system memory limit or was otherwise unable to handle the form.

@@

\subsection{On the correlation between operation count and performance}
As reiterated throughout the paper, the execution time of a kernel is influenced not only by the operation count, but also by the chosen code generation strategy (e.g., loop unrolling rather than preserving the loop structure). Figure~\ref{fig:flops-correlation} aggregates the \texttt{cfO2} results from the previous sections and displays them as a function of the reduction in executed flops. Besides summarizing the performance improvements achieved in this article, this figure highlights a number of aspects. The speed-ups achieved by \texttt{cfO2} tend to increase with the complexity of the operators and $q$. Further, the pre-multiplying functions have a limited impact in complex operators, and they occasionally have a negative effect in simple and small problems. Finally, this plot shows that even a powerful general-purpose compiler like Intel's is inadequate to fully exploit the structure of non-trivial finite element operators (analogous behaviors were observed with different compilers).

>>>>

We evaluate the speed-ups achieved by a wide variety of transformation systems over the ``original'' code produced by the FEniCS Form Compiler (i.e., no optimizations applied). We analyze the following transformation systems:
\begin{description}
\item[quad] Optimized quadrature mode. Work presented in~\citeN{quadrature1}, implemented in the FEniCS Form Compiler.
\item[tens] Tensor contraction mode. Work presented in~\citeN{FFC-TC}, implemented in the FEniCS Form Compiler.
\item[auto] Automatic choice between \texttt{tens} and \texttt{quad} driven by heuristic (detailed in~\citeN{Fenics} and summarized in Section~\ref{sec:qualitative}). Implemented in the FEniCS Form Compiler.
\item[ufls] UFLACS, a novel back-end for the FEniCS Form Compiler whose primary goals are improved code generation and execution times.
\item[cfO1] Generalized loop-invariant code motion. Work presented in~\citeN{Luporini}, implemented in COFFEE.
\item[cfO2] Optimal loop nest synthesis with handling of block-sparse tables. Work presented in this article, implemented in COFFEE.
\end{description}

@@

We report the results of our experiments in Figures~\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity} as three-dimensional plots. The axes represent $q$, $\mathrm{nf}$, and code transformation system. We show one subplot for each problem instance ${\langle} \textrm{form}, \mathrm{nf}, q {\rangle}$, with the code transformation system varying within each subplot. The best variant for each problem instance is given by the tallest bar, which indicates the maximum speed-up over the code produced by the FEniCS Form Compiler with no transformations applied (also referred to as ``baseline``). We note that if a bar or a subplot are missing, then the form compiler failed to generate code because it either exceeded the system memory limit or was otherwise unable to handle the form.

@@

\subsection{On the correlation between operation count and performance}
As reiterated throughout the paper, the execution time of a kernel is influenced not only by the operation count, but also by the chosen code generation strategy (e.g., loop unrolling rather than preserving the loop structure). Figure~\ref{fig:flops-correlation} aggregates the \texttt{cfO2} results from the previous sections and displays them as a function of the reduction in flops counted at the source level. Besides summarizing the performance improvements achieved in this article, this figure highlights a number of aspects. The speed-ups achieved by \texttt{cfO2} tend to increase with the complexity of the operators and $q$. Further, the pre-multiplying functions have a limited impact in complex operators. Finally, this plot shows that even a powerful general-purpose compiler like Intel's is inadequate to fully exploit the structure of non-trivial finite element operators (analogous behaviors were observed with different compilers).
